---
title: "dplyr_notes"
output: html_document
---

```{r setup}
library(dplyr)
library(ggplot2)
```
  
We will be practicing with `dplyr` functions using the `starwars` data set 

```{r}
starwars
head(starwars)
?starwars
```
  
###Subset rows with `filter`  
filter() allows us to subset the **rows** of a data set based on the values  

```{r}
## Filter for rows with human Star Wars characters
filter(starwars, species == "Human")

## save to a new R object
humans <- filter(starwars, species == "Humans")
humans
filter(humans, sex == "male")
```
  
**We can use different comparison operators to filter:**  
• `!` is "not"  
• `==` is "equal""  
• `&` is "and"  
• `|` is "or"  
• `<` is "less than"  
•  `>` is "greater than"  
• `is.na()` returns TRUE if value is NA 

```{r}
## return characters with a missing mass
filter(starwars, is.na(mass))
```
```{r}
## how to get all characters that are NOT human?
filter(starwars, species != "human")
```
  
variable_name `%in%` vector_of_values returns TRUE if value of variable_name is equal to one of the values in vector_of_values.  

This is a shortcut for variable_name == vector_of_values[1] | variable_name == vector_of_values[2]...  

```{r}
skin_color <- "grey"
skin_color %in% c("yellow", "fair", "tan") # returns false because "grey" is not in a vector of colors.

##How would we filter for characters with brown or blond hair?
filter(starwars, hair_color %in% c("blond", "brown"))
```
  
grepl("characters_of_interest", variable_name) return TRUE if the value of variable_ name contains the characters_of_interest.
```{r}
filter(starwars, grepl("blond", hair_color))
## How would we get all characters with blue in their eyes?
filter(starwars, grepl("blue", eye_color))
```
  
You can use `<`, `>`, `<=`, `>=` to compare numeric variables or categorical variables  
• nominal categorical variables (character vectors) are ranked alphabetically  
• ordinal categorical variables (factors) are ranked based on the built-in order  
```{r}
## starwars character with a name starting with letters before "D"
filter(starwars, name < "D") #ranking based on alphabetical order

str(diamonds)
filter(diamonds, cut > "Good") # ranking is based on levels of the ordered factor cut
"D" < "d" #false
"d" < "D" #true
```
  
between(x, left, right) is a shorthand for x >= left & x<= right  
```{r}
between(5,1,7) #true because 1 <= 5 <= 7
filter(starwars, between(mass, 50, 100))
```

## Practice!  
Find all starwars characters that  
1. Are missing a hair color
```{r}
filter(starwars, is.na(hair_color))
```
  
2. Are shorter than 100 cm and are not droids
```{r}
filter(starwars, height < 100 & species != "droid")
```
  
3. Were born in between the years 100 and 200  
```{r}
filter(starwars, between(birth_year, 100, 200))
```
  
4. Another of your choice!  
```{r}
filter(starwars, sex == "female")
```
  
## Ordering Rows with arrange  
arrange()reorders rows based on the values of column(s)  
```{r}
#order by character names
arrange(starwars, name)
# order of character values: aAbBcCdD

## order by species and birth year
arrange(starwars, species, birth_year)

##use desc() to get birth years in descending order
arrange(starwars, species, desc(birth_year))
```
  
# Practice!  
1. Arrange the starwars dataset from tallest to shortest characters
```{r}
arrange(starwars, desc(height))
```
  
2. Arrange the starwars dataset from shortest to tallest by species
```{r}
arrange(starwars, species, height) # if height and species switched, order by height tie breaker is alphabetic species
```

3. Alphabetize the Star Wars characters by their hair color then eye color.
```{r}
arrange(starwars, hair_color, eye_color)
```


## Subset columns with `select()`
```{r}
select(starwars, hair_color, skin_color, eye_color)
character_colors <- select(starwars, hair_color, skin_color, eye_color)
character_colors
```
  
We can also select multiple columns in a row  
```{r}
select(starwars, name:eye_color)
```


We can also remove column by name
```{r}
starwars
select(starwars, -films, -vehicles, -starships)
select(starwars, -c(films, vehicles, starships)) 
select(starwars, -c(films: starships)) # same as above
```
  
  
# Helpful functions  

`everything()` is a shortcut for "everything else"
```{r}
select(starwars, species, everything()) # put species first, then all the other columns

## alphabetize columns
sw_names <- sort(names(starwars))
sw_names
select(starwars, sw_names)
```


`starts_with("abc")` matches columns names that start with "abc"  
`ends_with("abc")` matches columns names that end with "abc"  
`contains("abc")` matches columns names that contain "abc"   

```{r}
select(starwars, starts_with("h"))

select(starwars, contains("or"))
```


# Practice! (5 min)
    
1. Select each character's homeworld and species. What question might this subset of variables answer?  
   
2.Brainstorm as many ways as possible to select `hair_color`, `skin_color`, `eye_color`, and `birth_year`

```{r}
#1
select(starwars, homeworld, species)
# We can observe the most common species that inhabit a world

#2
select(starwars, hair_color, skin_color, eye_color, birth_year)
select(starwars, c(hair_color, skin_color, eye_color, birth_year))
select(starwars, hair_color: birth_year)
select(starwars, c(hair_color: birth_year))
select(starwars, contains("_"))
select(starwars, contains("color"), birth_year)

```

   
## Add new variables with `mutate`  

```{r}
starwars_small <- select(starwars, height, mass, species)
starwars_small

mutate(starwars_small,
       height_m = height/100)
```

   
We can also create multiple variables at once and we can refer to variables we just created  
```{r}
mutate(starwars_small,
       height_m = height/100, 
       bmi = mass/height_m^2)
```

Summary statistics give one value for the entire column: `sum(), mean()`  
```{r}
mutate(starwars_small,
       sum = sum(mass, na.rm = T))
mutate(starwars_small,
       row_addition = height + mass,
       summary_sum = sum(height, mass, na.rm = T))
```

    
Group by a variable using the `.by` parameter  
```{r}
mutate(starwars_small,
       sum = sum(mass, na.rm = T),
       .by = species)
```

Change the type of variable  
```{r}
mutate(starwars_small,
       mass = as.character(mass))
```

## Practice (2 mins)
1. Convert mass to pounds (1kg ~ 2.22bs)  
```{r}
mutate(starwars_small,
       mass_lbs = mass*2.2)
```

## Summarize data with `summarize`  
```{r}
summarise(starwars,
          mean_mass = mean(mass, na.rm = T))
#Group by Species

summarise(starwars,
          mean_mass = mean(mass, na.rm = T),
          .by = species)
```

   
The only columns kept are the columns you group by and the columns you create.  

`n()` is a helpful function that returns the number of rows in a group

```{r}
summarise(starwars,
          mean_mass = mean(mass, na.rm = T),
          count = n(),
          .by = species)
##group by multiple groups
summarise(starwars,
          mean_mass = mean(mass, na.rm = T),
          count = n(),
          .by = c(species,sex))
```
  
## Renaming  
rename(data, new_name = old_name)
```{r}
new_starwars <- rename(starwars,
       birthYear = birth_year)

```


